High-Fidelity Counterfactual Explanations through Conformal Prediction

### 2023-05-04 notebooks/gmsc.qmd
work on tuning JEM

# Real-World Data

@@

# Data:
n_obs = 10000
counterfactual_data = load_california_housing(n_obs)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
output_dim = length(unique(labels))
```

@@

```{julia}
# Model parameters:
epochs = 100
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 64
activation = Flux.relu
builder = MLJFlux.@builder Flux.Chain(
    Dense(n_in, n_hidden, activation),
    Dense(n_hidden, n_out),
)
n_ens = 5                                   # number of models in ensemble
_loss = Flux.Losses.logitcrossentropy       # loss function
_finaliser = x -> x                         # finaliser function
```

@@

# Joint Energy Model:
jem = JointEnergyClassifier(
    sampler;
    builder=builder,
    epochs=epochs,
    batch_size=batch_size,
    finaliser=_finaliser,
    loss=_loss,
    jem_training_params=(
        α=α,verbosity=10,
    ),
    sampling_steps=20,
)

@@

```{julia}
# # Evaluate models:

# measure = Dict(
#     :f1score => multiclass_f1score, 
#     :acc => accuracy, 
#     :precision => multiclass_precision
# )
# model_performance = DataFrame()
# for (mod_name, mod) in model_dict
#     # Test performance:
#     test_data = load_mnist_test()
#     _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
#     _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
#     _perf.mod_name .= mod_name
#     model_performance = vcat(model_performance, _perf)
# end
# Serialization.serialize(joinpath(output_path,"gmsc_model_performance.jls"), model_performance)
# CSV.write(joinpath(output_path, "gmsc_model_performance.csv"), model_performance)
# model_performance
```

@@

bmk = benchmark(
    counterfactual_data; 
    models=model_dict, 
    generators=generator_dict, 
    measure=measures,
    suppress_training=true, dataname="Californian Housing",
    n_individuals=5,
    factual=0, target=1,
    initialization=:identity,
)
CSV.write(joinpath(output_path, "gmsc_benchmark.csv"), bmk())
```

@@

```{julia}
df = @chain bmk() begin
    @filter(variable in [
        "distance_from_energy",
        "distance_from_targets",
        "distance",])
    @mutate(variable = ifelse.(variable .== "distance_from_energy", "Non-Conformity", variable))
    @mutate(variable = ifelse.(variable .== "distance_from_targets", "Implausibility", variable))
    @mutate(variable = ifelse.(variable .== "distance", "Cost", variable))
end
plt = AlgebraOfGraphics.data(df) * visual(BoxPlot) * 
    mapping(:generator, :value, row=:variable, col=:model, color=:generator)
plt = draw(
    plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=150, height=120), 
    facet=(; linkyaxes=:minimal)
)   
display(plt)
save(joinpath(output_images_path, "gmsc_benchmark.png"), plt, px_per_unit=5)
```

>>>>

# GMSC

@@

# Data:
test_size = 0.2
n_obs = Int(1000 / (1.0 - test_size))
counterfactual_data, test_data = train_test_split(load_gmsc(n_obs); test_size=test_size)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
output_dim = length(unique(labels))
```

@@

```{julia}
# Model parameters:
epochs = 100
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 256
activation = Flux.relu
builder = MLJFlux.@builder Flux.Chain(
    Dense(n_in, n_hidden, activation),
    Dense(n_hidden, n_hidden, activation),
    Dense(n_hidden, n_hidden, activation),
    Dense(n_hidden, n_hidden, activation),
    Dense(n_hidden, n_out),
)
n_ens = 5                                   # number of models in ensemble
_loss = Flux.Losses.logitcrossentropy       # loss function
_finaliser = x -> x                         # finaliser function
```

@@

# Joint Energy Model:
jem = JointEnergyClassifier(
    sampler;
    builder=builder,
    epochs=epochs,
    batch_size=batch_size,
    finaliser=_finaliser,
    loss=_loss,
    jem_training_params=(
        α=α,verbosity=10,
    ),
    sampling_steps=60,
)

@@

```{julia}
# Evaluate models:

measure = Dict(
    :f1score => multiclass_f1score, 
    :acc => accuracy, 
    :precision => multiclass_precision
)
model_performance = DataFrame()
for (mod_name, mod) in model_dict
    # Test performance:
    _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
    _perf.mod_name .= mod_name
    model_performance = vcat(model_performance, _perf)
end
Serialization.serialize(joinpath(output_path,"gmsc_model_performance.jls"), model_performance)
CSV.write(joinpath(output_path, "gmsc_model_performance.csv"), model_performance)
model_performance
```

@@

bmk = benchmark(
    counterfactual_data; 
    models=model_dict, 
    generators=generator_dict, 
    measure=measures,
    suppress_training=true, dataname="California Housing",
    n_individuals=5,
    factual=0, target=1,
    initialization=:identity,
)
CSV.write(joinpath(output_path, "gmsc_benchmark.csv"), bmk())
```

@@

```{julia}
df = @chain bmk() begin
    @mutate(variable = ifelse.(variable .== "distance_from_energy", "Non-Conformity", variable))
    @mutate(variable = ifelse.(variable .== "distance_from_targets", "Implausibility", variable))
    @mutate(variable = ifelse.(variable .== "distance", "Cost", variable))
    @mutate(variable = ifelse.(variable .== "redundancy", "Redundancy", variable))
    @mutate(variable = ifelse.(variable .== "Validity", "Validity", variable))
end
plt = AlgebraOfGraphics.data(df) * visual(BoxPlot) * 
    mapping(:generator, :value, row=:variable, col=:model, color=:generator)
plt = draw(
    plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=150, height=120), 
    facet=(; linkyaxes=:none)
)   
display(plt)
save(joinpath(output_images_path, "gmsc_benchmark.png"), plt, px_per_unit=5)
```

### 2023-05-04 notebooks/mnist.qmd
work on tuning JEM

```{julia}
function pre_process(x; noise::Float32=1/256f0)
    ϵ = Float32.(randn(size(x)) * noise)
    x += ϵ
    return x
end
```

@@

# Data:
n_obs = 1000
counterfactual_data = load_mnist(n_obs)
counterfactual_data.X = pre_process.(counterfactual_data.X)
counterfactual_data.generative_model = vae
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
x_factual = reshape(pre_process(x_factual, noise=0.0f0), input_dim, 1)
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
# JEM parameters:
𝒟x = Uniform(0,1)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=10,
)
α = [1.0,1.0,5e-2]      # penalty strengths
```

@@

```{julia}
# Plot generated samples:
n_regen = 120
if _regen 
    for (mod_name, mod) in model_dict
        if ECCCo._has_sampler(mod)
            sampler = ECCCo._get_sampler(mod)
        else
            K = length(counterfactual_data.y_levels)
            input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
            𝒟x = Uniform(extrema(counterfactual_data.X)...)
            𝒟y = Categorical(ones(K) ./ K)
            sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size)
        end
        opt = ImproperSGLD()
        f(x) = logits(mod, x)

@@

```{julia}
function _plot_eccco_mnist(
    x::Union{AbstractArray, Int}=x_factual, target::Int=target;
    λ=[0.5,0.1,0.5],
    temp=0.1,η=0.01,
    plt_order = ["MLP", "MLP Ensemble", "JEM", "JEM Ensemble"],
    opt = Flux.Optimise.Adam(η),
    rng::Union{Int,AbstractRNG}=Random.GLOBAL_RNG,
)

@@

```{julia}
plt, eccco_generator = _plot_eccco_mnist()
display(plt)
savefig(plt, joinpath(output_images_path, "mnist_eccco.png"))
```

>>>>

```{julia}
function pre_process(x; noise::Float32=0.03f0)
    ϵ = Float32.(randn(size(x)) * noise)
    x += ϵ
    return x
end
```

@@

# Data:
n_obs = 10000
counterfactual_data = load_mnist(n_obs)
counterfactual_data.X = pre_process.(counterfactual_data.X)
counterfactual_data.generative_model = vae
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
x_factual = reshape(pre_process(x_factual, noise=0.0f0), input_dim, 1)
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
# JEM parameters:
𝒟x = Uniform(0,1)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=10,
)
α = [1.0,1.0,1e-1]      # penalty strengths
```

@@

```{julia}
# Plot generated samples:
n_regen = 200
if _regen 
    for (mod_name, mod) in model_dict
        K = length(counterfactual_data.y_levels)
        input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
        𝒟x = Uniform(extrema(counterfactual_data.X)...)
        𝒟y = Categorical(ones(K) ./ K)
        sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size, prob_buffer=0.0)
        opt = ImproperSGLD()
        f(x) = logits(mod, x)

@@

```{julia}
function _plot_eccco_mnist(
    x::Union{AbstractArray, Int}=x_factual, target::Int=target;
    λ=[0.1,0.1,0.1],
    temp=0.1,η=0.01,
    plt_order = ["MLP", "MLP Ensemble", "JEM", "JEM Ensemble"],
    opt = Flux.Optimise.Adam(η),
    rng::Union{Int,AbstractRNG}=Random.GLOBAL_RNG,
    T::Int = 100,
)

@@

```{julia}
plt, eccco_generator = _plot_eccco_mnist(opt=JSMADescent(η=1.0))
display(plt)
savefig(plt, joinpath(output_images_path, "mnist_eccco.png"))
```

### 2023-05-04 notebooks/moons.qmd
work on tuning JEM

# Circles Data

@@

# Data:
test_size = 0.2
n_obs = Int(1000 / (1.0 - test_size))
counterfactual_data, test_data = train_test_split(load_circles(n_obs); test_size=test_size)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
output_dim = length(unique(labels))
```

@@

```{julia}
# Model parameters:
epochs = 100
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 32
activation = Flux.swish
builder = MLJFlux.MLP(
    hidden=(n_hidden, n_hidden, n_hidden,), 
    σ=activation
)
n_ens = 5                                   # number of models in ensemble
_loss = Flux.Losses.logitcrossentropy       # loss function
_finaliser = x -> x                         # finaliser function
```

```{julia}
# JEM parameters:
𝒟x = Normal()
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=batch_size,
)
α = [1.0,1.0,1e-2]      # penalty strengths
```

@@

```{julia}
n_regen = 256
n_each = batch_size
for (mod_name, mod) in model_dict
    K = length(counterfactual_data.y_levels)
    input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
    𝒟x = Uniform(extrema(counterfactual_data.X)...)
    𝒟y = Categorical(ones(K) ./ K)
    sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size, prob_buffer=0.0)
    opt = ImproperSGLD()
    plts = []
    for target in levels(labels)
        X̂ = sampler(f, opt; niter=n_regen, n_samples=n_each, y=target)
        ex = extrema(hcat(MLJFlux.reformat(X),X̂), dims=2)
        xlims = ex[1]
        ylims = ex[2]
        x1 = range(1.0f0.*xlims...,length=100)
        x2 = range(1.0f0.*ylims...,length=100)
        target_idx = findall(levels(labels) .== target)[1]
        f(x) = target_idx == 2 ? probs(mod, x) : 1 .- probs(mod, x)
        plt = Plots.contour(
            x1, x2, (x, y) -> f([x, y][:,:])[1], 
            fill=true, alpha=0.5, title="Target: $target", cbar=true,
            xlims=xlims,
            ylims=ylims,
        )
        Plots.scatter!(
            MLJFlux.reformat(X)[1,:], MLJFlux.reformat(X)[2,:], 
            color=Int.(labels.refs).-1, group=Int.(labels.refs).-1, alpha=0.5
        )
        Plots.scatter!(
            X̂[1,:], X̂[2,:], 
            color=repeat([target], size(X̂,2)), 
            group=repeat([target], size(X̂,2)), 
            shape=:star5, ms=10
        )
        savefig(plt, joinpath(output_images_path, "circles_generated_$(mod_name).png"))
        push!(plts, plt)
    end
    plt = Plots.plot(plts..., layout=(1, 2), size=(2*500, 400), plot_title=mod_name)
    display(plt)
end
```

>>>>

# Moons Data

@@

# Data:
test_size = 0.2
n_obs = Int(1000 / (1.0 - test_size))
counterfactual_data, test_data = train_test_split(load_moons(n_obs); test_size=test_size)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
output_dim = length(unique(labels))
```

@@

```{julia}
# Model parameters:
epochs = 100
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 128
activation = Flux.relu
builder = MLJFlux.@builder Flux.Chain(
    Dense(n_in, n_hidden, activation),
    Dense(n_hidden, n_hidden, activation),
    Dense(n_hidden, n_out),
)
n_ens = 5                                   # number of models in ensemble
_loss = Flux.Losses.logitcrossentropy       # loss function
_finaliser = x -> x                         # finaliser function
```

```{julia}
# JEM parameters:
𝒟x = Normal()
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=10,
)
α = [1.0,1.0,1e-1]      # penalty strengths
```

@@
